A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-10-25 JNABRev20231018/ComparingNLStools.Rmd
minor updates

For R **developers** we invite and encourage discussion of the design choices, since these
have downstream implications for ease of use, adaptation to new features and 
efficiency of ongoing maintenance.

@@

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
task that seems straightforward but is quite nasty. Its very succinct statement 
provides the "short reproducible example" much requested on R mailing lists. 
This is a real problem from a field researcher. The data and graph follow.

@@

Our emphasis on the software resilience to starting parameters that are, in the
sense of statistical modeling, "silly" comes from over half a century of dealing
with users whose interests and understanding are very far from those of statistical
modelers. Thus we seek methods and codes that obtain reasonable answers under
highly unfavourable conditions. Nevertheless, a proper approach to 
nonlinear modeling is to apply all available knowledge to the task.
<!-- end 231018 work -->

@@

```{r ex02set, echo=TRUE}
# model formulas
frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
#
# Starting parameter sets
stu1<-c(b1=1, b2=1, b3=1)
sts1<-c(c1=1, c2=1, c3=1)
stt1<-c(Asym=1, xmid=1, scal=1)
```

@@

```{r ex02fn, echo=TRUE}
# Logistic3U
hobbs.res  <-  function(x){ # unscaled Hobbs weeds problem -- residual
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
  # Note: this form of the residual, from Laplace (1788) in the form
  # of "fitted - observed" has been used in our software for half
  # a century, but sometimes offends users of the more traditional
  # "observed - fitted" form
}

hobbs.jac  <-  function(x) { # unscaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-x[3]*tt)
  zz  <-  1.0/(1+x[2]*yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -x[1]*zz*zz*yy
  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
  attr(jj, "gradient") <- jj
  jj
}
```

@@

```{r ex05, echo=TRUE}  
stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
print(badstart)
```

@@

The design goals of the different tools may also be revealed in the so-called
"convergence tests" for the iterative solvers. In the manual page for `nls()` 
in R 4.0.0 there was the warning:

@@

As mentioned, the output of `nls()`, `minpack.lm::nlsLM()`, or
`gslnls::gsl_nls()` is an object of class "nls" which has a quite rich structure
described in the manual files or revealed by applying the `str()` function to the
result of `nls()`.  The complexity of this object is a challenge to users. Let us 
use for example `result <- snlm1` as the returned object from `nlsLM()` for the 
Logistic3S problem. . The `data` return element 
is an R symbol. To actually access the data from this element, we need to use 
the syntax:
```
eval(parse(text=result$data))
```
However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

@@

Tools that produce a class `nls` output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
`nlsr` computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian 
are actually computed in the `print` and `summary` methods for the result. 
These two approaches
lead to different consequences for performance and how features are
provided. `nlsr` has antecedents in the methods of @cnm79, where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in `nlsr` is likely of value for workers to copy and modify
for customized tools.

@@

- Karl Schilling (private communication) provided an example where a model specified
  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
  the derivative is undefined, but the model can be computed. In such cases, 
  we observed that `nls()` and 
  `minpack.lm::nlsLM` found a solution, though this seems to be a lucky accident.

@@

- `nls()` can only impose bounds if the `algorithm="port"` argument is 
used in the call. Unfortunately, the documentation warns us:

@@

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
             upper=c(2,6,3), algorithm='port')
pnls(anlshob1b) #  check the answer (short form)
# nlsLM seems NOT to work with bounds in this example
anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls(anlsLM1b)
# also no warning if starting out of bounds, but gets a good answer!!
st4<-c(c1=4, c2=4, c3=4)
anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pshort(anlx1b)
```

@@

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of the problem
require(nlsr); require(minpack.lm); traceval<-FALSE
stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, 3), trace=traceval))
pnls(anlM1)
anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
pnls(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,30))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
             upper=c(2,6,30), algorithm='port')
pnls(anlshmsk1)
anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                 upper=c(2,6,30))
pnls(anlsLMmsks1)

# Test with all parameters masked
anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
print(anlxmskall)
```

@@

An important choice made in developing `nlsr` was to code entirely within the R 
programming language. `nls()` uses a mix of R, C, and Fortran, as does `minpack.lm`. 
`gslnls` is an R wrapper to various C-language routines in the Gnu Scientific 
Library (@GSL-manual).
Generally, keeping to a single programming language can allow for 
easier maintenance and upgrades. It also avoids some work when there are changes 
or upgrades to libraries for the non-R languages. 
R is usually considered slower 
than most computing environments because it keeps track of objects and 
because it is usually interpreted. In recent years, the 
performance penalty for using code entirely in R has been much reduced
with the just-in-time compiler and other improvements. All-R computation 
may now offer acceptable performance. In `nlsr`, the use of R may be a
smaller performance cost than the aggressive approach to a solution,
which can cause more iterations to be used.

@@

nlsModel <- function(form, data, start, wts, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
{
    ## thisEnv <- environment() # shared by all functions in the 'm' list; variable no longer needed
    env <- new.env(hash = TRUE, parent = environment(form))
    for(i in names(data)) env[[i]] <- data[[i]]
    ind <- as.list(start)
    parLength <- 0L
    for(i in names(ind) ) {
        temp <- start[[i]]
        storage.mode(temp) <- "double"
        env[[i]] <- temp
        ind[[i]] <- parLength + seq_along(temp)
        parLength <- parLength + length(temp)
    }
    getPars.noVarying <- function() unlist(mget(names(ind), env))
    getPars <- getPars.noVarying
    internalPars <- getPars()

@@

```{r nlswtx, echo=TRUE}
wts <- 0.5^tt # simple weights
frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
Asym<-1; xmid<-1; scal<-1
nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
attr(rnowt, "gradient")<-NULL; rnowt
usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
attr(rusewt,"gradient")<-NULL; rusewt
## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0,"gradient")<-NULL; rn0 # weighted residuals at starting coefficients
nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
rn<-nmod$resid()
attr(rn,"gradient")<-NULL; rn # same as rusewt
```

@@

The variable projection method (@Golub1973, @OlearyRust13) is 
usually much more effective than general approaches in finding good solutions 
to nonlinear least squares 
problems when some of the parameters appear linearly. In our logistic examples, 
the asymptote parameters are an illustration. However, identifying which parameters
are linear and communicating this information to estimating functions 
is not a trivial task. `nls()` has an option `algorithm="plinear"` that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability. The `nlstac` package uses a different algorithm for similar goals.

@@

Such goals align with the aims of **unit testing** 
(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
and the conventional R package testing tools).
In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarization of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

>>>>

For R **developers**, we invite and encourage discussion of the design choices, since these
have downstream implications for ease of use, adaptation to new features, and 
efficiency of ongoing maintenance.

@@

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth-curve modeling 
task that seems straightforward but is quite nasty. Its very succinct statement 
provides the "short reproducible example" much requested on R mailing lists. 
This is a real problem from a field researcher. The data and graph follow.

@@

Our emphasis on the software resilience to starting parameters that are, in the
sense of statistical modeling, "silly" comes from over half a century of dealing
with users whose interests and understanding are very far from those of statistical
modelers. Thus, we seek methods and codes that obtain reasonable answers under
highly unfavourable conditions. Nevertheless, a proper approach to 
nonlinear modeling is to apply all available knowledge to the task.
<!-- end 231018 work -->

@@

```{r ex02set, echo=TRUE}
# model formulas
frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
frms <- weed ~ 100 * c1/(1 + 10* c2* exp(-0.1 * c3* tt))
frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
#
# Starting parameter sets
stu1 <- c(b1 = 1, b2 = 1, b3 = 1)
sts1 <- c(c1 = 1, c2 = 1, c3 = 1)
stt1 <- c(Asym = 1, xmid = 1, scal = 1)
```

@@

```{r ex02fn, echo=TRUE}
# Logistic3U
hobbs.res <- function(x){ # unscaled Hobbs weeds problem -- residual
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt <- 1:12
  res <- x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
  # Note: this form of the residual, from Laplace (1788) in the form
  # of "fitted - observed" has been used in our software for half
  # a century, but sometimes offends users of the more traditional
  # "observed - fitted" form
}

hobbs.jac <- function(x) { # unscaled Hobbs weeds problem -- Jacobian
  jj <- matrix(0.0, 12, 3)
  tt <- 1:12
  yy <- exp(-x[3] * tt)
  zz <- 1.0 / (1 + x[2] * yy)
  jj[tt, 1] <- zz
  jj[tt, 2] <- -x[1] * zz * zz * yy
  jj[tt, 3] <- x[1] * zz * zz * yy * x[2] * tt
  attr(jj, "gradient") <- jj
  jj
}
```

@@

```{r ex05, echo=TRUE}  
stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
badstart <- nlxb(formula = frmt, start = stspecial, data = weeddf)
print(badstart)
```

@@

The design goals of the different tools may also be revealed in the so-called
"convergence tests" for the iterative solvers. In the manual page for `nls()` 
in R 4.0.0, there was the warning:

@@

As mentioned, the output of `nls()`, `minpack.lm::nlsLM()`, or
`gslnls::gsl_nls()` is an object of class "nls" which has a quite rich structure
described in the manual files or revealed by applying the `str()` function to the
result of `nls()`.  The complexity of this object is a challenge to users. Let us 
use for example `result <- snlm1` as the returned object from `nlsLM()` for the 
Logistic3S problem. The `data` return element 
is an R symbol. To actually access the data from this element, we need to use 
the syntax:
```
eval(parse(text = result$data))
```
However, if the call is made with `model = TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

@@

Tools that produce a class `nls` output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
`nlsr` computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian 
are actually computed in the `print` and `summary` methods for the result. 
These two approaches
lead to different consequences for performance and how features are
provided. `nlsr` has antecedents in the methods of @cnm79, where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus, the code in `nlsr` is likely of value for workers to copy and modify
for customized tools.

@@

- Karl Schilling (private communication) provided an example where a model specified
  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
  derivative w.r.t. `b` is `a * (x ^ b * log(x))`. If there is data for which `x = 0`,
  the derivative is undefined, but the model can be computed. In such cases, 
  we observed that `nls()` and 
  `minpack.lm::nlsLM` found a solution, though this seems to be a lucky accident.

@@

- `nls()` can only impose bounds if the `algorithm = "port"` argument is 
used in the call. Unfortunately, the documentation warns us:

@@

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
             upper = c(2,6,3), algorithm = 'port')
pnls(anlshob1b) #  check the answer (short form)
# nlsLM seems NOT to work with bounds in this example
anlsLM1b <- nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
pnls(anlsLM1b)
# also no warning if starting out of bounds, but gets a good answer!!
st4 <- c(c1 = 4, c2 = 4, c3 = 4)
anlsLMob <- nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
pnls(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
pshort(anlx1b)
```

@@

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of the problem
require(nlsr); require(minpack.lm); traceval <- FALSE
stu <- c(b1 = 200, b2 = 50, b3 = 0.3) # a default starting vector (named!)
sts <- c(c1 = 2, c2 = 5, c3 = 3) # a default scaled starting vector (named!)
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
			upper = c(200, 60, 3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
			upper=c(200, 60, 3), trace = traceval))
pnls(anlM1)
anlsmsk1 <- try(nls(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
		upper = c(200, 60, 3),  algorithm = "port", trace = traceval))
pnls(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start = sts, data = weeddf, lower = c(2, 0, 0),
                  upper = c(2, 6, 30))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start = sts, trace = traceval, data = weeddf, 
                 lower = c(2, 0, 0), upper = c(2, 6, 30), algorithm = 'port')
pnls(anlshmsk1)
anlsLMmsks1 <- nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
                 upper = c(2,6,30))
pnls(anlsLMmsks1)

# Test with all parameters masked
anlxmskall <- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
print(anlxmskall)
```

@@

An important choice made in developing `nlsr` was to code entirely within the R 
programming language. `nls()` uses a mix of R, C, and Fortran, as does `minpack.lm`. 
`gslnls` is an R wrapper to various C-language routines in the GNU Scientific 
Library (@GSL-manual).
Generally, keeping to a single programming language can allow for 
easier maintenance and upgrades. It also avoids some work when there are changes 
or upgrades to libraries for the non-R languages. 
R is usually considered slower 
than most computing environments because it keeps track of objects and 
because it is usually interpreted. In recent years, the 
performance penalty for using code entirely in R has been much reduced
with the just-in-time compiler and other improvements. All-R computation 
may now offer acceptable performance. In `nlsr`, the use of R may be a
smaller performance cost than the aggressive approach to a solution,
which can cause more iterations to be used.

@@

nlsModel <- function(form, data, start, wts, upper = NULL, scaleOffset = 0,
                     nDcentral = FALSE)
{
    ## thisEnv <- environment() # shared by all functions in the 'm' list; variable no longer needed
    env <- new.env(hash = TRUE, parent = environment(form))
    for(i in names(data)) env[[i]] <- data[[i]]
    ind <- as.list(start)
    parLength <- 0L
    for(i in names(ind)) {
        temp <- start[[i]]
        storage.mode(temp) <- "double"
        env[[i]] <- temp
        ind[[i]] <- parLength + seq_along(temp)
        parLength <- parLength + length(temp)
    }
    getPars.noVarying <- function() unlist(mget(names(ind), env))
    getPars <- getPars.noVarying
    internalPars <- getPars()

@@

```{r nlswtx, echo=TRUE}
wts <- 0.5 ^ tt # simple weights
frmlogis <- weed ~ Asym / (1 + exp((xmid - tt)/scal))
Asym <- 1; xmid <- 1; scal <- 1
nowt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
rnowt <- nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
attr(rnowt, "gradient") <- NULL; rnowt
usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights = wts)
rusewt <- usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
attr(rusewt, "gradient") <- NULL; rusewt
## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
nmod0 <- nlsModel(frmlogis, data = weeddf, start = c(Asym = 1, xmid = 1, scal = 1), wts = wts)
rn0 <- nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0, "gradient") <- NULL; rn0 # weighted residuals at starting coefficients
nmod <- nlsModel(frmlogis, data = weeddf, start = coef(usewt), wts = wts)
rn <- nmod$resid()
attr(rn,"gradient")<-NULL; rn # same as rusewt
```

@@

The variable projection method (@Golub1973, @OlearyRust13) is 
usually much more effective than general approaches in finding good solutions 
to nonlinear least squares 
problems when some of the parameters appear linearly. In our logistic examples, 
the asymptote parameters are an illustration. However, identifying which parameters
are linear and communicating this information to estimating functions 
is not a trivial task. `nls()` has an option `algorithm = "plinear"` that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability. The `nlstac` package uses a different algorithm for similar goals.

@@

Such goals align with the aims of **unit testing** 
(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
and the conventional R package testing tools).
In our work, one of us (AB) has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarization of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.
